\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2015-02-09 doc/flexiblesusy-paper.tex
fixing typos found by Jae-hyeon Park

None of the fundamental motivations of supersymmetry require
minimality, and specific alternatives to (or extensions of) the MSSM
are, for example, motivated by the $\mu$-problem of the MSSM
\cite{Kim:1983dt}; explaining the family structure (see
e.g.~\cite{King:2014nza}) or for successful baryogenesis or
leptogenesis (see e.g.~\cite{King:2008qb}). However constructing
specialized tools to study all relevant models would require an
enormous amount of work.  So general tools which can automate this
process and produce fast and reliable programs can greatly enhance our
ability to understand and test non-minimal realizations of
supersymmetry.

@@

Exploring the parameter space of supersymmetric models with a high
number of free parameters is quite time consuming.  Therefore \fs aims
to produce spectrum generators with a short run-time.  The two most
time consuming parts of a SUSY spectrum generator are usually the
calculation of the two-loop $\beta$-functions and the pole masses of
mixed particles:
%
\begin{itemize}
\item \emph{Calculation of the $\beta$-functions:} The RG solving
  algorithms usually need $O(10)$ iterations between the high and the
  low scale to find a set of parameters consistent will all boundary
  conditions with a $0.01\%$ precision goal.  During each iteration
  the Runge-Kutta algorithm needs to calculate all $\beta$-functions
  $O(50)$ times.  Most two-loop $\beta$-functions involve $O(50)$
  matrix multiplications and additions.  All together one arrives at
  $O(10^4)$ matrix operations.  To optimize these, \fs uses the fast
  linear algebra package \href{Eigen}{http://eigen.tuxfamily.org}.
  Eigen uses C++ expression templates to remove temporary objects and
  enable lazy evaluation of the expressions.  It supports explicit
  vectorization, and provides fixed-size matrices to avoid dynamic
  memory allocation.  All of these features in combination result in
  very fast code for the calculation of the $\beta$-functions in \fs.
%
\item \emph{Calculation of the pole masses:} The second most time
  consuming part is the precise calculation of the pole masses for
  mixed particles.  For each particle $\psi_k$ in a multiplet the full
  self-energy matrix $\Sigma^\psi_{ij}(p=m^\text{tree}_{\psi_k})$ has
  to be evaluated.  Each self-energy matrix entry again involves the
  calculation of $O(50)$ Feynman diagrams, each involving the
  calculation of vertices and a loop-function.  All in all, one
  arrives at $O(500)$ Feynman diagrams and $O(10^4)$ loop function
  evaluations.  To speed up the calculation of the pole masses \fs
  makes use of multi-threading, where each pole mass is calculated in
  a separate thread.  This allows the operating system to distribute
  these calculations among different CPU cores.  With this technique
  one can gain a speed-up of $20$--$30\%$.
\end{itemize}

@@

The model boundary conditions are defined in the \fs model file
\code{FlexibleSUSY.m}, which has to be located in the model directory
\code{models/<model>/}.  To add this the user should create a
\code{FlexibleSUSY.m.in} file in the directory
\code{model_files/<model>/}.  When the \code{./createmodel} script is
executed, the \code{FlexibleSUSY.m} file is created from the
\code{model_files/<model-file-name>/FlexibleSUSY.m.in} file, where the
directory \code{<model-file-name>} is the specified by the
\code{--model-file=<model-file-name>} option.  If no such option is
given the directory matching the \code{--name=<model>} option is used.
In either case the \code{FlexibleSUSY.m} file which is created is then
automatically placed in the directory \code{models/<model>/}.  Note
that many predefined example model files can already be found in
\code{model_files/}.

@@

The variables \code{LowScaleInput}, \code{SUSYScaleInput}

@@

 The corrections can then be used in the calculation of the Higgs
 masses, when appropriate settings are selected in the SLHA file. Note
 that even in the NMSSM the user must make an important decision as
 to whether or not to enable the generated MSSM corrections which are
 incomplete in the NMSSM.  We feel that it is valuable to have
 these MSSM corrections for scenarios where singlet mixing is
 very small and in particular for cross checks against the MSSM when
 close to the MSSM limit of the model.  However in cases where the
 singlet mixing is large the result at $O(y_t^4 + y_t^2 y_b^2 +
 y_b^4)$ and $O(y_\tau^4)$ will not be complete and including these
 corrections could in principle even make the numerical result further
 away from the correct two loop result at that order if there is a
 cancellation with the missing contributions.  So while including such
 partial two-loop corrections does not change the formal accuracy
 the user should choose whether or not to include these corrections
 based on the physics they study.  Similarly in models that go beyond
 the NMSSM the user must decide based the physics whether on not these
 corrections will give the leading two-loop corrections in their
 model.

@@

Finally, with the variable \code{ExtraSLHAOutputBlocks} the user can
define extra SLHA output blocks.  The values of the block entries will
be calculated at the output scale, which is specified in entry $12$ in
the SLHA input block \code{MODSEL}.  In the example model file above
the following six extra SLHA compliant output blocks are defined:
\code{ALPHA}, \code{HMIX}, \code{Au}, \code{Ad}, \code{Ae} and
\code{MSOFT}.  The \code{ALPHA} output block contains the CP-even
Higgs pole mass mixing angle as the only entry.  The \code{HMIX}
output block contains the $\mu$-parameter, the ratio $\tan\beta =
v_u/v_d$, the combination $v=\sqrt{v_u^2 + v_d^2}$, the squared mass
of the CP-odd Higgs $m_A^2$, the soft-breaking parameter $B\mu$ and
the values of $v_u$ and $v_d$, all defined in the \DRbar\ scheme.  In
an analogous way four more output blocks for the soft-breaking \DRbar\
parameters are defined.  For a CMSSM example parameter point with $m_0
= 125\unit{GeV}$, $M_{1/2} = 500\unit{GeV}$, $\tan\beta = 10$,
$\sign\mu = +1$ and $A_0 = 0$ the \fs-generated CMSSM spectrum
generator writes the so defined blocks to the output in the form
%
\begin{lstlisting}
Block ALPHA Q= 8.76740936E+02
           1.06784138E-01   # ArcCos(Pole(ZH(0,1)))
Block HMIX Q= 8.76740936E+02
     1     6.31218384E+02   # Mu
     2     9.67312621E+00   # vu/vd
     3     2.44053433E+02   # Sqrt(Sqr(vd) + Sqr(vu))
     4     5.36777230E+05   # Sqr(MAh(1))
   101     5.49048159E+04   # BMu
   102     2.50962986E+01   # vd
   103     2.42759663E+02   # vu
Block Au Q= 8.76740936E+02
  1  1    -1.14477419E+03   # TYu(0,0)/Yu(0,0)
  2  2    -1.14476911E+03   # TYu(1,1)/Yu(1,1)
  3  3    -8.83902977E+02   # TYu(2,2)/Yu(2,2)
Block Ad Q= 8.76740936E+02
  1  1    -1.40026447E+03   # TYd(0,0)/Yd(0,0)
  2  2    -1.40025976E+03   # TYd(1,1)/Yd(1,1)
  3  3    -1.30885006E+03   # TYd(2,2)/Yd(2,2)
Block Ae Q= 8.76740936E+02
  1  1    -3.00005426E+02   # TYe(0,0)/Ye(0,0)
  2  2    -3.00000006E+02   # TYe(1,1)/Ye(1,1)
  3  3    -2.98364373E+02   # TYe(2,2)/Ye(2,2)
Block MSOFT Q= 8.76740936E+02
     1     2.09018579E+02   # MassB
     2     3.88257873E+02   # MassWB
     3     1.11544211E+03   # MassG
    21     1.09683411E+05   # mHd2
    22    -3.85898988E+05   # mHu2
    31     3.54416224E+02   # Sqrt(ml2(0,0))
    32     3.54412891E+02   # Sqrt(ml2(1,1))
    33     3.53407333E+02   # Sqrt(ml2(2,2))
    34     2.22035720E+02   # Sqrt(me2(0,0))
    35     2.22024873E+02   # Sqrt(me2(1,1))
    36     2.18731935E+02   # Sqrt(me2(2,2))
    41     1.02053367E+03   # Sqrt(mq2(0,0))
    42     1.02053107E+03   # Sqrt(mq2(1,1))
    43     9.40760849E+02   # Sqrt(mq2(2,2))
    44     9.82565930E+02   # Sqrt(mu2(0,0))
    45     9.82563167E+02   # Sqrt(mu2(1,1))
    46     8.09126982E+02   # Sqrt(mu2(2,2))
    47     9.77979392E+02   # Sqrt(md2(0,0))
    48     9.77976666E+02   # Sqrt(md2(1,1))
    49     9.73121951E+02   # Sqrt(md2(2,2))
\end{lstlisting}

@@

All model parameters and their $\beta$-functions are contained in
the first and second derived classes.  The structure of the
$\beta$-functions of a general supersymmetric model
\cite{Jones:1974pg,Jones:1983vk,West:1984dg,Martin:1993yx,Yamada:1993ga,Jack:1994kd,Jack:1994rk,Yam94,MV94,Fonseca:2011vn,Goodsell:2012fm,Sperling:2013eva,Sperling:2013xqa}
allows to split these parameters into two classes:
%
\begin{enumerate}
\item \emph{SUSY parameters:} gauge couplings, superpotential
  parameters and VEVs and
\item \emph{soft-breaking parameters} \cite{Girardello:1981wz}: soft
  linear scalar terms, soft bilinear scalar interactions, soft
  trilinear scalar interactions, soft gaugino mass terms and soft
  scalar squared masses.
\end{enumerate}
%
The $\beta$-functions of the SUSY parameters in general depend only on
the SUSY parameters and are independent of the soft-breaking
parameters.  However, the $\beta$-functions of the soft-breaking
parameters depend on all model parameters in general.
%
This property is reflected in the C++ code: The class
\code{<model>_susy_parameters} directly inherits from
\code{Beta_functions} and implements the $\beta$-functions of the SUSY
parameters.  The class of soft-breaking parameters
\code{<model>_soft_parameters} in turn inherits from
\code{<model>_susy_parameters} and implements the $\beta$-functions of
the soft-breaking parameters in terms of all model parameters.  The so
constructed class hierarchy allows to (i) use the RGE running of all
model parameters via the common \code{Beta_function} interface and to
(ii) run the SUSY parameters independently of the soft-breaking
parameters.

@@

\fs assumes that each SUSY model contains Higgs bosons, which trigger
a spontaneous breaking of the electroweak symmetry.  The corresponding
EWSB consistency conditions are formulated in \fs at the one-loop
level as
%
\begin{align}
  0 &= \frac{\partial V^\text{tree}}{\partial v_i} - t_i,
  & & (i=1,\ldots,N)
  \label{eq:one-loop-ewsb-eq}
\end{align}
%
where $V^\text{tree}$ is the tree-level Higgs potential, $v_i$ is the
VEV corresponding to the Higgs fields $H_i$ and $t_i$ is the one-loop
tadpole diagram of $H_i$.  Already at the tree-level ($t_i = 0$) Eqs.\
\eqref{eq:one-loop-ewsb-eq} can have multiple solutions, depending on
which parameters are chosen to be fixed by these equations.  A
well-known example is the real MSSM, where the $\mu$-parameter is
chosen to be fixed by the EWSB equations.  In this model, the EWSB
equations, Eqs.\ \eqref{eq:one-loop-ewsb-eq} can only determine
$|\mu|$, while the sign of $\mu$ is not fixed.  This results in two
solutions for $\mu$ of the form
%
\begin{align}
  \mu = \sign\mu \cdot |\mu| .
  \label{eq:EWSB-solution-for-mu}
\end{align}
%
Currently, \fs handles only real model parameters and renormalizable
theories, which restricts the number of possible multiple solutions of
Eqs.\ \eqref{eq:one-loop-ewsb-eq} to be less or equal to $4N$.  In
some cases, as for example in the real MSSM, the different solutions
are related by one or more global signs for the parameters, as for example in
Eq.~\eqref{eq:EWSB-solution-for-mu}.  \fs recognizes such cases and
introduces these global signs as additional free parameters to allow
the user to choose between the different solutions.  The case of the
real MSSM Eq.~\eqref{eq:EWSB-solution-for-mu} is therefore handled
automatically in \fs, because $\sign\mu$ is automatically introduced
as an additional free parameter.

@@

It is important to note that in the case of such problem points it is
non-trivial to judge whether this is because there is no physical
solution for the given parameter space point or whether a solution
exists but the fixed point iteration is unable to find the solution.
While \fs makes it as easy as possible to find spectra, when studying
new models a physical understanding of the model is still essential
and this can help the user determine why the such problems arise.

@@

\url{http://iktp.tu-dresden.de/~avoigt/FlexibleSUSY/test.xhtml}

@@

One of \fs's design goals is a short run-time.  In this section we
demonstrate that this goal was achieved by comparing the run-time of
two different sets of CMSSM spectrum generators:
%
\begin{itemize}
\item \emph{Without sfermion flavour violation:} Disallowing sfermion
  flavour violation simplifies the calculation of the pole masses,
  because flavour-off-diagonal sfermion self-energy matrix elements
  do not need to be calculated.  Here we compare \fs's non-flavour
  violating CMSSM spectrum generator FlexibleSUSY-NoFV (version 1.0.0)
  against SPheno (version 3.2.4) and Softsusy (version 3.4.0).
%
\item \emph{With sfermion flavour violation:} Allowing for sfermion
  flavour violation in general increases the run-time of spectrum
  generators, because the full $6\times 6$ sfermion self-energy
  matrices have to be calculated.  Here we compare FlexibleSUSY-FV
  (version 1.0.0) and SPhenoMSSM (generated with SARAH 4.1.0 and
  linked against SPheno 3.2.4).  Both spectrum generators are based on
  SARAH's MSSM model file, which allows for sfermion flavour
  violation.
\end{itemize}
%
For the run-time comparison the CKM matrix is set to unity, all
$CP$-violating phases are set to zero and $R$-parity violation is
disabled.  \fs and Softsusy are compiled with g++ 4.8.0 and Intel
ifort 13.1.3 20130607.  SPheno and SPhenoMSSM are compiled with Intel
ifort 13.1.3 20130607.\footnote{Intel's ifort compiler decreases the
  run-time of SPheno and SPhenoMSSM by approximately a factor $1.5$,
  compared to gfortran.}  We are generating $2\cdot 10^{4}$ random
CMSSM parameter points with $m_0\in [50,1000]\unit{GeV}$, $m_{1/2}\in
[50,1000]\unit{GeV}$, $\tan\beta\in [1,100]$, $\sign\mu\in \{-1,+1\}$
and $A_0\in [-1000,$ $1000]\unit{GeV}$.  For each point an SLHA input
file is created by appending the values of $m_0$, $m_{1/2}$,
$\tan\beta$, $\sign\mu$, $A_0$ in form of a \code{MINPAR} block to the
SLHA template file given in \ref{sec:speed-test-slha-template-file}.
The resulting SLHA input file is passed to each spectrum generator and
the (wall-clock) time is measured until the program has finished.  The
average run-times for three different CPU types can be found in
\tabref{tab:run-time-comparison}.  The first column shows the run-time
on a Intel Core2 Duo (P8600, $2.40\unit{GHz}$) where only one core was
enabled.  The second column shows the run-time on the same processor
where both cores were enabled.  In the third column a machine with two
Intel Xeon CPUs (L5640, $2.27\unit{GHz}$, $6$ cores) was used.
%
\begin{table}[tbh]
  \centering
  \begin{tabular}{llll}
    \toprule
                            & Intel Core2 Duo    & Intel Core2 Duo   & $2$ $\times$ Intel Xeon\\
                            & (P8600, $1$ core)  & (P8600, $2$ cores)& (L5640, $6$ cores)\\
    \midrule
    FlexibleSUSY-NoFV 1.0.0 & $0.086\unit{s}$    & $0.079\unit{s}$   & $0.060\unit{s}$\\
    SPheno 3.2.4            & $0.119\unit{s}$    & $0.114\unit{s}$   & $0.101\unit{s}$\\
    Softsusy 3.4.0          & $0.175\unit{s}$    & $0.171\unit{s}$   & $0.147\unit{s}$\\
    \midrule
    FlexibleSUSY-FV 1.0.0   & $0.150\unit{s}$    & $0.113\unit{s}$   & $0.074\unit{s}$\\
    SPhenoMSSM 4.1.0        & $0.415\unit{s}$    & $0.401\unit{s}$   & $0.370\unit{s}$\\
    \bottomrule
  \end{tabular}
  \caption{Average run-time of CMSSM spectrum generators
    for random parameter points.  The first three rows show three
    spectrum generators with disabled sfermion flavour violation.  Rows
    $4$--$5$ show two spectrum generators with enabled sfermion flavour violation, both based
    on SARAH's MSSM model file.}
  \label{tab:run-time-comparison}
\end{table}

Under both the non-flavour violating spectrum generators (first three
rows) as well as the flavour violating ones (4th and 5th row) we find
that \fs is significantly fastest.  Compared to SPheno,
FlexibleSUSY-NoFV is faster by a factor $1.4$--$1.7$, and compared to
Softsusy around a factor $2$--$2.5$.  Under the flavour violating
spectrum generators FlexibleSUSY-FV is faster than SPhenoMSSM by a
factor $2.8$--$5$.  Reason for the long run-time of SPhenoMSSM is the
long calculation duration of the two-loop $\beta$-functions.  Here \fs
benefits a lot from Eigen's well-optimizable matrix expressions.  We
also find that increasing the number of CPU cores reduces the run-time
of \fs.  The reason is that \fs calculates each pole mass in a
separate thread, and therefore benefits from multi-core CPUs.

>>>>

None of the fundamental motivations of supersymmetry requires
minimality, and specific alternatives to (or extensions of) the MSSM
are, for example, motivated by the $\mu$-problem of the MSSM
\cite{Kim:1983dt}; explaining the family structure (see
e.g.~\cite{King:2014nza}) or for successful baryogenesis or
leptogenesis (see e.g.~\cite{King:2008qb}). However constructing
specialized tools to study all relevant models would require an
enormous amount of work.  So general tools which can automate this
process and produce fast and reliable programs can greatly enhance our
ability to understand and test non-minimal realizations of
supersymmetry.

@@

Exploring the parameter space of supersymmetric models with a high
number of free parameters is quite time consuming.  Therefore \fs aims
to produce spectrum generators with a short run-time.  The two most
time consuming parts of a SUSY spectrum generator are usually the
calculation of the two-loop $\beta$-functions and the pole masses of
mixed particles:
%
\begin{itemize}
\item \emph{Calculation of the $\beta$-functions:} The RG solving
  algorithms usually need $O(10)$ iterations between the high and the
  low scale to find a set of parameters consistent with all boundary
  conditions with a $0.01\%$ precision goal.  During each iteration
  the Runge-Kutta algorithm needs to calculate all $\beta$-functions
  $O(50)$ times.  Most two-loop $\beta$-functions involve $O(50)$
  matrix multiplications and additions.  All together one arrives at
  $O(10^4)$ matrix operations.  To optimize these, \fs uses the fast
  linear algebra package \href{Eigen}{http://eigen.tuxfamily.org}.
  Eigen uses C++ expression templates to remove temporary objects and
  enable lazy evaluation of the expressions.  It supports explicit
  vectorization, and provides fixed-size matrices to avoid dynamic
  memory allocation.  All of these features in combination result in
  very fast code for the calculation of the $\beta$-functions in \fs.
%
\item \emph{Calculation of the pole masses:} The second most time
  consuming part is the precise calculation of the pole masses for
  mixed particles.  For each particle $\psi_k$ in a multiplet the full
  self-energy matrix $\Sigma^\psi_{ij}(p=m^\text{tree}_{\psi_k})$ has
  to be evaluated.  Each self-energy matrix entry again involves the
  calculation of $O(50)$ Feynman diagrams, each involving the
  calculation of vertices and a loop-function.  All in all, one
  arrives at $O(500)$ Feynman diagrams and $O(10^4)$ loop function
  evaluations.  To speed up the calculation of the pole masses \fs
  makes use of multi-threading, where each pole mass is calculated in
  a separate thread.  This allows the operating system to distribute
  these calculations among different CPU cores.  With this technique
  one can gain a speed-up of $20$--$30\%$.
\end{itemize}

@@

The model boundary conditions are defined in the \fs model file
\code{FlexibleSUSY.m}, which has to be located in the model directory
\code{models/<model>/}.  To add this the user should create a
\code{FlexibleSUSY.m.in} file in the directory
\code{model_files/<model>/}.  When the \code{./createmodel} script is
executed, the \code{FlexibleSUSY.m} file is created from the
\code{model_files/<model-file-name>/FlexibleSUSY.m.in} file, where the
directory \code{<model-file-name>} is specified by the
\code{--model-file=<model-file-name>} option.  If no such option is
given the directory matching the \code{--name=<model>} option is used.
In either case the \code{FlexibleSUSY.m} file which is created is then
automatically placed in the directory \code{models/<model>/}.  Note
that many predefined example model files can already be found in
\code{model_files/}.

@@

The variables \code{LowScaleInput}, \code{SUSYScaleInput} and

@@

 The corrections can then be used in the calculation of the Higgs
 masses, when appropriate settings are selected in the SLHA file. Note
 that even in the NMSSM the user must make an important decision as
 to whether or not to enable the generated MSSM corrections which are
 incomplete in the NMSSM.  We feel that it is valuable to have
 these MSSM corrections for scenarios where singlet mixing is
 very small and in particular for cross checks against the MSSM when
 close to the MSSM limit of the model.  However in cases where the
 singlet mixing is large the result at $O(y_t^4 + y_t^2 y_b^2 +
 y_b^4)$ and $O(y_\tau^4)$ will not be complete and including these
 corrections could in principle even make the numerical result further
 away from the correct two loop result at that order if there is a
 cancellation with the missing contributions.  So while including such
 partial two-loop corrections does not change the formal accuracy
 the user should choose whether or not to include these corrections
 based on the physics they study.  Similarly in models that go beyond
 the NMSSM the user must decide based on the physics whether on not these
 corrections will give the leading two-loop corrections in their
 model.

@@

Finally, with the variable \code{ExtraSLHAOutputBlocks} the user can
define extra SLHA output blocks.  The values of the block entries will
be calculated at the output scale, which is specified in entry $12$ in
the SLHA input block \code{MODSEL}.  In the example model file above
the following six extra SLHA compliant output blocks are defined:
\code{ALPHA}, \code{HMIX}, \code{Au}, \code{Ad}, \code{Ae} and
\code{MSOFT}.  The \code{ALPHA} output block contains the $CP$-even
Higgs pole mass mixing angle as the only entry.  The \code{HMIX}
output block contains the $\mu$-parameter, the ratio $\tan\beta =
v_u/v_d$, the combination $v=\sqrt{v_u^2 + v_d^2}$, the squared mass
of the $CP$-odd Higgs $m_A^2$, the soft-breaking parameter $B\mu$ and
the values of $v_u$ and $v_d$, all defined in the \DRbar\ scheme.  In
an analogous way four more output blocks for the soft-breaking \DRbar\
parameters are defined.  For a CMSSM example parameter point with $m_0
= 125\unit{GeV}$, $M_{1/2} = 500\unit{GeV}$, $\tan\beta = 10$,
$\sign\mu = +1$ and $A_0 = 0$ the \fs-generated CMSSM spectrum
generator writes the so defined blocks to the output in the form
%
\begin{lstlisting}
Block ALPHA Q= 8.76740936E+02
           1.06784138E-01   # ArcCos(Pole(ZH(0,1)))
Block HMIX Q= 8.76740936E+02
     1     6.31218384E+02   # Mu
     2     9.67312621E+00   # vu/vd
     3     2.44053433E+02   # Sqrt(Sqr(vd) + Sqr(vu))
     4     5.36777230E+05   # Sqr(MAh(1))
   101     5.49048159E+04   # BMu
   102     2.50962986E+01   # vd
   103     2.42759663E+02   # vu
Block Au Q= 8.76740936E+02
  1  1    -1.14477419E+03   # TYu(0,0)/Yu(0,0)
  2  2    -1.14476911E+03   # TYu(1,1)/Yu(1,1)
  3  3    -8.83902977E+02   # TYu(2,2)/Yu(2,2)
Block Ad Q= 8.76740936E+02
  1  1    -1.40026447E+03   # TYd(0,0)/Yd(0,0)
  2  2    -1.40025976E+03   # TYd(1,1)/Yd(1,1)
  3  3    -1.30885006E+03   # TYd(2,2)/Yd(2,2)
Block Ae Q= 8.76740936E+02
  1  1    -3.00005426E+02   # TYe(0,0)/Ye(0,0)
  2  2    -3.00000006E+02   # TYe(1,1)/Ye(1,1)
  3  3    -2.98364373E+02   # TYe(2,2)/Ye(2,2)
Block MSOFT Q= 8.76740936E+02
     1     2.09018579E+02   # MassB
     2     3.88257873E+02   # MassWB
     3     1.11544211E+03   # MassG
    21     1.09683411E+05   # mHd2
    22    -3.85898988E+05   # mHu2
    31     3.54416224E+02   # Sqrt(ml2(0,0))
    32     3.54412891E+02   # Sqrt(ml2(1,1))
    33     3.53407333E+02   # Sqrt(ml2(2,2))
    34     2.22035720E+02   # Sqrt(me2(0,0))
    35     2.22024873E+02   # Sqrt(me2(1,1))
    36     2.18731935E+02   # Sqrt(me2(2,2))
    41     1.02053367E+03   # Sqrt(mq2(0,0))
    42     1.02053107E+03   # Sqrt(mq2(1,1))
    43     9.40760849E+02   # Sqrt(mq2(2,2))
    44     9.82565930E+02   # Sqrt(mu2(0,0))
    45     9.82563167E+02   # Sqrt(mu2(1,1))
    46     8.09126982E+02   # Sqrt(mu2(2,2))
    47     9.77979392E+02   # Sqrt(md2(0,0))
    48     9.77976666E+02   # Sqrt(md2(1,1))
    49     9.73121951E+02   # Sqrt(md2(2,2))
\end{lstlisting}

@@

All model parameters and their $\beta$-functions are contained in
the first and second derived classes.  The structure of the
$\beta$-functions of a general supersymmetric model
\cite{Jones:1974pg,Jones:1983vk,West:1984dg,Martin:1993yx,Yamada:1993ga,Jack:1994kd,Jack:1994rk,Yam94,MV94,Fonseca:2011vn,Goodsell:2012fm,Sperling:2013eva,Sperling:2013xqa}
allows to split these parameters into two classes:
%
\begin{enumerate}
\item \emph{SUSY parameters:} gauge couplings, superpotential
  parameters and VEVs and
\item \emph{soft-breaking parameters} \cite{Girardello:1981wz}: soft
  linear scalar terms, soft bilinear scalar interactions, soft
  trilinear scalar interactions, soft quadlinear scalar interactions,
  soft gaugino mass terms and soft
  scalar squared masses.
\end{enumerate}
%
The $\beta$-functions of the SUSY parameters in general depend only on
the SUSY parameters and are independent of the soft-breaking
parameters.  However, the $\beta$-functions of the soft-breaking
parameters depend on all model parameters in general.
%
This property is reflected in the C++ code: The class
\code{<model>_susy_parameters} directly inherits from
\code{Beta_functions} and implements the $\beta$-functions of the SUSY
parameters.  The class of soft-breaking parameters
\code{<model>_soft_parameters} in turn inherits from
\code{<model>_susy_parameters} and implements the $\beta$-functions of
the soft-breaking parameters in terms of all model parameters.  The so
constructed class hierarchy allows to (i) use the RGE running of all
model parameters via the common \code{Beta_function} interface and to
(ii) run the SUSY parameters independently of the soft-breaking
parameters.

@@

\fs assumes that each SUSY model contains Higgs bosons, which trigger
a spontaneous breaking of the electroweak symmetry.  The corresponding
EWSB consistency conditions are formulated in \fs at the one-loop
level as
%
\begin{align}
  0 &= \frac{\partial V^\text{tree}}{\partial v_i} - t_i,
  & & (i=1,\ldots,N)
  \label{eq:one-loop-ewsb-eq}
\end{align}
%
where $V^\text{tree}$ is the tree-level Higgs potential, $v_i$ is the
VEV corresponding to the Higgs field $H_i$ and $t_i$ is the one-loop
tadpole diagram of $H_i$.  Already at the tree-level ($t_i = 0$) Eqs.\
\eqref{eq:one-loop-ewsb-eq} can have multiple solutions, depending on
which parameters are chosen to be fixed by these equations.  A
well-known example is the real MSSM, where the $\mu$-parameter is
chosen to be fixed by the EWSB equations.  In this model, the EWSB
equations, Eqs.\ \eqref{eq:one-loop-ewsb-eq} can only determine
$|\mu|$, while the sign of $\mu$ is not fixed.  This results in two
solutions for $\mu$ of the form
%
\begin{align}
  \mu = \sign\mu \cdot |\mu| .
  \label{eq:EWSB-solution-for-mu}
\end{align}
%
Currently, \fs handles only real model parameters and renormalizable
theories, which restricts the number of possible multiple solutions of
Eqs.\ \eqref{eq:one-loop-ewsb-eq} to be less than or equal to $4N$.  In
some cases, as for example in the real MSSM, the different solutions
are related by one or more global signs for the parameters, as for example in
Eq.~\eqref{eq:EWSB-solution-for-mu}.  \fs recognizes such cases and
introduces these global signs as additional free parameters to allow
the user to choose between the different solutions.  The case of the
real MSSM Eq.~\eqref{eq:EWSB-solution-for-mu} is therefore handled
automatically in \fs, because $\sign\mu$ is automatically introduced
as an additional free parameter.

@@

It is important to note that in the case of such problem points it is
non-trivial to judge whether this is because there is no physical
solution for the given parameter space point or a solution
exists but the fixed point iteration is unable to find the solution.
While \fs makes it as easy as possible to find spectra, when studying
new models a physical understanding of the model is still essential
and this can help the user determine why such problems arise.

@@

\url{https://www.desy.de/~alvoigt/FlexibleSUSY/test.xhtml}

@@

One of \fs's design goals is a short run-time.  In this section we
demonstrate that this goal was achieved by comparing the run-time of
two different sets of CMSSM spectrum generators:
%
\begin{itemize}
\item \emph{Without sfermion flavour violation:} Disallowing sfermion
  flavour violation simplifies the calculation of the pole masses,
  because flavour-off-diagonal sfermion self-energy matrix elements
  do not need to be calculated.  Here we compare \fs's non-flavour
  violating CMSSM spectrum generator FlexibleSUSY-NoFV (version 1.0.0)
  against SPheno (version 3.2.4) and Softsusy (version 3.4.0).
%
\item \emph{With sfermion flavour violation:} Allowing for sfermion
  flavour violation in general increases the run-time of spectrum
  generators, because the full $6\times 6$ sfermion self-energy
  matrices have to be calculated.  Here we compare FlexibleSUSY-FV
  (version 1.0.0) and SPhenoMSSM (generated with SARAH 4.1.0 and
  linked against SPheno 3.2.4).  Both spectrum generators are based on
  SARAH's MSSM model file, which allows for sfermion flavour
  violation.
\end{itemize}
%
For the run-time comparison the CKM matrix is set to unity, all
$CP$-violating phases are set to zero and $R$-parity violation is
disabled.  \fs and Softsusy are compiled with g++ 4.8.0 and Intel
ifort 13.1.3 20130607.  SPheno and SPhenoMSSM are compiled with Intel
ifort 13.1.3 20130607.\footnote{Intel's ifort compiler decreases the
  run-time of SPheno and SPhenoMSSM by approximately a factor $1.5$,
  compared to gfortran.}  We are generating $2\cdot 10^{4}$ random
CMSSM parameter points with $m_0\in [50,1000]\unit{GeV}$, $m_{1/2}\in
[50,1000]\unit{GeV}$, $\tan\beta\in [1,100]$, $\sign\mu\in \{-1,+1\}$
and $A_0\in [-1000,$ $1000]\unit{GeV}$.  For each point an SLHA input
file is created by appending the values of $m_0$, $m_{1/2}$,
$\tan\beta$, $\sign\mu$, $A_0$ in form of a \code{MINPAR} block to the
SLHA template file given in \ref{sec:speed-test-slha-template-file}.
The resulting SLHA input file is passed to each spectrum generator and
the (wall-clock) time is measured until the program has finished.  The
average run-times for three different CPU types can be found in
\tabref{tab:run-time-comparison}.  The first column shows the run-time
on an Intel Core2 Duo (P8600, $2.40\unit{GHz}$) where only one core was
enabled.  The second column shows the run-time on the same processor
where both cores were enabled.  In the third column a machine with two
Intel Xeon CPUs (L5640, $2.27\unit{GHz}$, $6$ cores) was used.
%
\begin{table}[tbh]
  \centering
  \begin{tabular}{llll}
    \toprule
                            & Intel Core2 Duo    & Intel Core2 Duo   & $2$ $\times$ Intel Xeon\\
                            & (P8600, $1$ core)  & (P8600, $2$ cores)& (L5640, $6$ cores)\\
    \midrule
    FlexibleSUSY-NoFV 1.0.0 & $0.086\unit{s}$    & $0.079\unit{s}$   & $0.060\unit{s}$\\
    SPheno 3.2.4            & $0.119\unit{s}$    & $0.114\unit{s}$   & $0.101\unit{s}$\\
    Softsusy 3.4.0          & $0.175\unit{s}$    & $0.171\unit{s}$   & $0.147\unit{s}$\\
    \midrule
    FlexibleSUSY-FV 1.0.0   & $0.150\unit{s}$    & $0.113\unit{s}$   & $0.074\unit{s}$\\
    SPhenoMSSM 4.1.0        & $0.415\unit{s}$    & $0.401\unit{s}$   & $0.370\unit{s}$\\
    \bottomrule
  \end{tabular}
  \caption{Average run-time of CMSSM spectrum generators
    for random parameter points.  The first three rows show three
    spectrum generators with disabled sfermion flavour violation.  Rows
    $4$--$5$ show two spectrum generators with enabled sfermion flavour violation, both based
    on SARAH's MSSM model file.}
  \label{tab:run-time-comparison}
\end{table}

Among both the non-flavour violating spectrum generators (first three
rows) as well as the flavour violating ones (4th and 5th row) we find
that \fs is significantly fastest.  Compared to SPheno,
FlexibleSUSY-NoFV is faster by a factor $1.4$--$1.7$, and compared to
Softsusy around a factor $2$--$2.5$.  Between the flavour violating
spectrum generators FlexibleSUSY-FV is faster than SPhenoMSSM by a
factor $2.8$--$5$.  Reason for the long run-time of SPhenoMSSM is the
long calculation duration of the two-loop $\beta$-functions.  Here \fs
benefits a lot from Eigen's well-optimizable matrix expressions.  We
also find that increasing the number of CPU cores reduces the run-time
of \fs.  The reason is that \fs calculates each pole mass in a
separate thread, and therefore benefits from multi-core CPUs.
